---
title: "Diamonds - An Exploratory Analysis"
author: "Luc Frachon"
date: "13 janvier 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This mini-project is another problem set from the Data Analysis in R course,part of the Udacity Data Analyst Nanodegree. It uses the diamonds dataset that ships with the ggplot2 package.

## Load data

```{r}
require(ggplot2)
data("diamonds")
str(diamonds)
summary(diamonds)
```

This dataset contains 53,940 observations, ie. data about 53,940 different diamonds including prices. We see that the data seems very clean as there as are no NAs. There are 0s in the x, y, z dimensions, however, which most likely correspond to missing values.

## Question 1: Scatterplot of price vs x

From the dataset's help file, we learn that x is the length in millimetres.

```{r}
ggplot(data = diamonds, aes(x, price)) +
    theme_light() +
    geom_point(alpha = .05, colour = 'steelblue')
```

The observations where x = 0 compress the horizontal scale. We can zoom in a little:

```{r}
ggplot(data = diamonds, aes(x, price)) +
    theme_light() +
    geom_point(alpha = .05, colour = 'steelblue') +
    scale_x_continuous(limits = c(3, 11), breaks = seq(3, 11, 2))
```

**Observations:**

 - There is a correlation between price and x that looks like an exponential
 - There are few stripes along some values of x, maybe indicating customary sizes in the diamond industry
 - The bulk of diamonds are between 3.5 and 7.0mm and between $300 and $7500, approximately

## Question 2: Correlations

 * **What is the correlation between price and x?**
 
```{r}
with(diamonds, cor(x, price, method = 'pearson'))
```

 * **What is the correlation between price and y?**
 
```{r}
with(diamonds, cor(y, price, method = 'pearson'))
```

 * **What is the correlation between price and z?**
 
```{r}
with(diamonds, cor(z, price, method = 'pearson'))
```

There are strong correlations, but they might be even higher if we transformed price by applying a log transformation (because of the exponential shape of the scatter plot). Let's try it for x:

```{r}
with(diamonds, cor(x, log10(price), method = 'pearson'))
```

As predicted, we have an almost perfect linear correlation.

## Question 3: Price vs. Depth

```{r}
ggplot(data = diamonds, aes(depth, price)) +
    theme_light() +
    geom_point(alpha = .05, colour = 'steelblue')
```

Again, we have a number of outliers but the majority of the observations is between 55 and 70 millmetres. Let us focus on that range:

```{r}
ggplot(data = diamonds, aes(depth, price)) +
    theme_light() +
    geom_point(alpha = .05, colour = 'steelblue') +
    coord_cartesian(xlim = c(55, 70))
```

With some adjustments to transparency and tick marks along the x-axis:

```{r}
ggplot(data = diamonds, aes(depth, price)) +
    theme_light() +
    geom_point(alpha = .01, colour = 'steelblue') +
    scale_x_continuous(breaks = seq(42, 80, 2)) +
    coord_cartesian(xlim = c(55, 70))
```

**Based on this scatterplot, most diamonds are between what values of depth?**

Most are between 59 and 64mm.

**Correlation of depth vs. price:**

```{r}
with(diamonds, cor(depth, price))
```

The correlation coefficient is very small and from the scatter plots, there is no reason to suspect some hidden, non-linear correlation.

## Question 4: Price vs. Carat

**Create a scatterplot of price vs carat and omit the top 1% of price and carat values.**

```{r}
dia_99pct <- subset(diamonds, 
                    price < .99 * max(price) & carat < .99 * max(carat))
ggplot(aes(carat, price), data = dia_99pct) +
    geom_point(alpha = 0.02, colour = 'steelblue')
```

## Question 5: Price vs. Volume

** Create a scatterplot of price vs. volume (x * y * z). This is a very rough approximation for a diamond's volume. Create a new variable for volume in the diamonds data frame. This will be useful in a later exercise. Don't make any adjustments to the plot just yet.**

```{r}
diamonds$volume <- with(diamonds, x * y * z)
qplot(data = diamonds, volume, price)
```


